```{r, include = FALSE}
source("global_stuff.R")
```

# Behaviorism

```{r, cache=FALSE}
word_count <- wordcountaddin::word_count("C6_Behaviorism.Rmd")
wpm <- 200
reading_time <- paste0(round(word_count/wpm, 1), " minutes")
info <- data.frame(word_count,reading_time)
colnames(info) <- c("Word Count","Reading Time")
knitr::kable(info)
```

## Overview

Last chapter introduced associationism as an example of early ideas about how processes of cognition might work. This chapter covers the school of behaviorism, which rose to prominence in American psychology in the early 20th century (roughly between 1920s to 1960s). Behaviorism embraced some aspects of associationism, but took a decidedly applied approach to the study of cognitive abilities. Cognitive psychology is sometimes described as a revolution against the school of behaviorism, but the transition was very slow, and core claims and goals of behaviorism continue to have implications for the cognitive sciences and society in general today.

## The Rabbit Hole: To explain or not to explain?

One of the goals of this textbook is to examine how empirical evidence is used to test and develop explanations about how cognition works. In my view, the task of explaining how cognition works is an important goal of research into cognitive abilities. Why is explanation important? I think there are several reasons. Tentative theories provide guidance and clues about how to focus research inquiries, and good explanations can lead to the development of useful applications and technologies. I also think that explanations are intrinsically interesting and worthwhile for understanding ourselves and our relationship to the world around us. At the same time, not everyone agrees that explanation is intrinsically important or useful. 

For the moment, let's consider a fictional future where the science of cognition has been fully developed in great detail, with rock-solid explanations about how human and animal cognition works. Presumably, the theories and findings would have useful applications. Perhaps, the theories would explain new ways to help people restore lost cognitive abilities or expand upon existing cognitive abilities, or make machines capable of cognitive abilities. Although the basic science of cognition is interested in explaining cognition for reasons beside possible applications, it is clear that the ability to manipulate and control cognitive abilities is likely to emerge from a mature cognitive science capable of explaining how cognitive abilities work. Indeed, in the fictional future, it is easy to imagine that future societies maintained funding of cognitive science because of its high potential to discover methods to manipulate and control cognition, that could lead to new applied fields like cognitive engineering and technology. It is also easy to imagine that tools to manipulate and control cognition could be scary, especially if they were used for nefarious purposes. Finally, as some forms of behaviorism argued, it is possible to develop methods for manipulating and controlling cognition and behavior in the absence of theory and explanation.

### Youtube, machine-learning, and modern behaviorism

The basic ideas of behaviorism have not disappeared, and they are being used in modern society in extraordinary ways to control and manipulate cognition and behavior. Let's start with a modern example to motivate a closer inspection of the history of behaviorism and its implications for understanding how cognition works. 

The  example is from a podcast called the [Rabbit Hole, by New York Times columnist Kevin Roose](https://www.nytimes.com/column/rabbit-hole). The series is generally about "what the internet is doing do us?". The first story in the podcast is a haunting account of how a young man's life was influenced in major ways by watching Youtube videos over a period of several years. How could watching youtube videos have such a major impact on someone's life? The answer lies in the powerful behavior changing methods being deployed on the internet by many companies. 

Youtube presents video content that users access through a web-browser. Web-browsers allow people to interact with the internet in many ways, like searching for content or clicking links on a webpage. Web-browsers also allow web-sites to embed code to harvest data about how people behave when they are on a website. For example, youtube can store the history of videos that you watched, the order of videos that you clicked on, the amount of time spent watching each video, and many other kinds of user data (e.g., your location, your comment history, etc.). Furthermore, the behavioral data collected by websites can be combined with statistical techniques (like [machine learning](https://en.wikipedia.org/wiki/Machine_learning)) to generate predictions about future user behavior.

Youtube collected massive amounts of video watching behavior data from their users, and then used advanced machine learning algorithms to predict and ultimately control and manipulate user behavior. For example, one goal was to improve the video recommendation algorithm, which seems like a useful goal for users of youtube. Improving the recommendation algorithm would help users find content they wanted to watch--very practical. Youtube also had the goal of increasing the amount of time that people spent watching youtube videos. If people watched more videos, then they would see more advertisements, and that would help youtube's bottom line. Youtube used the website collected behavioral data to generate predictions about what new videos to recommend to people in order to maximize their viewing time, and keep them watching youtube videos as long as possible.

For some youtube users, like the man in the podcast, the recommendation algorithm has deeply impacted their belief systems and behavior. For example, this user spent hours per day watching youtube videos, many of which were selected by the recommendation algorithm. Initially, his viewing habits included searches for quirky music videos, but over time the algorithm suggested new content that increased his viewing time. This new content happened to politically polarizing content that became increasingly extreme on the far-right of the political spectrum. The user credits watching youtube videos with his life decisions to become interested in and affiliated with white nationalist movements. Interestingly, the youtube algorithm may also be involved in taking him in and out of this rabbit hole. For example, after becoming involved in far-right movements he continued to watch youtube, which continued to suggest videos that would increase his viewing time. At this point in his viewing history, he became slightly more engaged by videos on the left-side of the political spectrum, and credits watching those videos with his decision to cease his involvement in far-right politics.

Although the cognitive sciences have been around for quite some time, there is no widely agreed upon theory or explanation of cognition that can be used to systematically control and manipulate behavior. Nevertheless, youtube's algorithms that predict future viewing behavior from large databases of past viewing behavior appear highly successful in manipulating video watching times. In addition to the above anecdote, in 2017, the Wall Street Journal reported youtube hit a new milestone--1 billion hours of youtube had been watched in a single day ^[that's 41.6 million days worth of videos, or 114,155 years worth of video]. Viewing time had increased ten times from 2012, thanks to the introduction of predictive algorithms credited with driving the behavioral change. At the close of this chapter, we will return to the youtube example to examine their algorithm as a modern case of implementing behavioral principles at massive scale. Next, we will examine the behaviorist approach to asking and answering questions about cognition.

## Watson's Behaviorism

Associationism is sometimes tied together with the school of [Behaviorism](https://en.wikipedia.org/wiki/Behaviorism) in psychology. Behaviorism came to prominence in American psychology between the 1910s and 1950s, and cognitive psychology is also portrayed somewhat as a reaction to behaviorism. Two prominent behaviorists were [John B. Watson](https://en.wikipedia.org/wiki/John_B._Watson) and [B.F. Skinner](https://en.wikipedia.org/wiki/B._F._Skinner). A basic tenet of behaviorism was that psychology should only use objective and observable measurements, and not use subjective measures, like introspection. On this view, questions about subjective experiences like mental imagery were not valid because they would rely on subjective introspective report. Instead, psychology should focus on measuring observable properties of stimuli and behavior, and should attempt to discover laws relating patterns of stimulation to patterns of behavioral responses.

Behaviorism has a complicated history beyond the scope of this book. However, it is worth making a few notes. First, behaviorism appeared alongside with the rise of experimental psychology as a new academic and scientific discipline in America. Early experimental psychologists were very concerned that psychology should be considered as a "hard natural science", and behaviorism partly reflects a desire among some psychologists to present themselves as hard scientists, and to make broad claims about what gets to count as science.

Second, behaviorism became popular alongside the eugenics movement. At first glance, it might seem that behaviorism and eugenics would be unrelated. For example, J. B. Watson, who wrote the "Behaviorist Manifesto" in 19XX, is famous for the following quote:

## Skinner's Radical Behaviorism

## Associationism and Behaviorism

## Tolman's Rats


