```{r, include = FALSE}
source("global_stuff.R")
```

# Information Processing

```{r, cache=FALSE}
word_count <- wordcountaddin::word_count("C7_Information_Processing.Rmd")
wpm <- 200
reading_time <- paste0(round(word_count/wpm, 1), " minutes")
info <- data.frame(word_count,reading_time)
colnames(info) <- c("Word Count","Reading Time")
knitr::kable(info)
```

## Information Theory

### Hick-Hyman Law

### Miller's Magic number 7

### What kind of theory is information theory?




we will discuss Pavlov's discovery of classical conditioning, and 


In contrast to the individual differences approach


Contrast theories of learning in terms of ones that do not assume memory, and ones that do.

- Memory-less learning theory...associative learning
- instance-based learning

Feats of learning

How do you learn something better or more efficiently?
How do you learn to a high degree of skill?

misconceptions about learning

Principles in skill learning

specificity
practice

<!--


### Associationism


Early experimental work in the 1890s by Edward Thorndike

is beyond the scope of this chapter, and


1890s...styles of explanation (emergent from simpler processes to claims of special psychic powers)...very new empirical territory...physiology of the brain...animal cognition.


## Animal Cognition

Instead of asking a question about whether a person is better or worse than someone else in relation to some task, consider questions like, how is cognition even possible? How do people form ideas, how do they come to know things, how do they understand and experience the world around them? More specifically, what are the psychological processes that enable all cognition, and how do they work?

Philosophers were among the people interested in these kinds of questions well before psychologists started asking similar questions. Unfortunately, a history of philosophy is outside the scope for this chapter [if you are interested, see]. Nevertheless, I will point to two philosophical issues that preceded and inspired psychological theory, dualism and associationism.

I will use dualism to refer to non-physicalist accounts of cognition. On these accounts, the mind somehow transcends the physical plane and can not be explained in terms of physical bodies and brains. A characteristic feature of dualism is the assignment of "special" qualities to aspects of cognition. For example, I could claim that the mind is special and can't be explained by the laws of physical reality. Note I did not provide any explanation of mind at all, I just made the claim that the mind requires "special sauce". It is worth keeping an eye out for "special sauce" claims, they pop up fairly often in cognition. The possibility that some aspect of cognition is special is not a problem; however, claiming that something is special without attempting to explain how the special thing works provides little explanatory value.

Physicalist accounts of cognition are much more constrained because the primary assumption is that cognition is a physical phenomena. There is no special (non-physical) sauce that can be invoked to explain cognition. Instead, human and non-human animal cognition is assumed to emerge from physical bodies and brains that interact with a physical environment. With these constraints in place, physicalist theorists of cognition are often interested in describing the algorithms behind cognition, and how they may be physically instantiated.

Associationist Philosophers are a good example of theorists who took a physicalist position, and attempted to explain aspects of cognition in terms of algorithmic processes, specifically, in terms of associations. In 1689, John Locke wrote, "An essay concerning human understanding", and argued against the nativist idea that people were born only with innate knowledge about the world. Instead, Locke advocated that people acquired knowledge by learning about the world through their experiences. David Hume elaborated on the nature of the learning process and suggested in, "Treatise on Human Nature", that "when the mind, therefore, passes from the idea or impression of one object to the idea or belief of another, it is not determined by reason, but by certain principles, which associate together the ideas of these objects, and unite them in the imagination."

If we fast-forward into the 1900s, we find that Experimental Psychologists embraced associationism and developed its ideas theoretically and empirically. Today, there are many theories of associative learning that attempt answers to Hume's conjecture. Hume referred to, "certain principles" that cause ideas to be associated together. What are the certain principles? What causes ideas to be associated together? The domain of associative learning has developed theories of associative learning processes, and used a back-and-forth process to update and improve theory with new data. 

In the next sections, we look at examples of basic claims made by an associationist theory of cognition, and methods used to gather evidence for testing the claims.
-->






